---
title: "GRSE tracking data setup"
author: "Luis Huckstadt"
date: "2025-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, Load libraries and setup}
library(aniMotum)
library(tidyverse)
library(readxl)
library(rnaturalearth)
library(availability)

deploy_tbl <- read_excel('../data/GESE Deployments 2024-2025.xlsx')|>
  janitor::clean_names() |>
  mutate(deployid = name) 
saveRDS(deploy_tbl,file = here::here('data/deploy_tbl.rds'))


# Plot tracks
## simple helper to plot tracks
plotTracks <- function(..., pal = cpal) {
  require(maptools)
  data(wrld_simpl)
  trks <- list(...)
  xlim <- do.call(range, lapply(trks, function(trk) trk[, 1]))
  ylim <- do.call(range, lapply(trks, function(trk) trk[, 2]))
  plot(xlim, ylim, type = "n", xlim = xlim, ylim = ylim, xlab = "longitude", ylab = "latitude")
  for (k in seq(from = floor((min(xlim)+180)/360), to = floor((max(xlim)+180)/360), by = 1))
      plot(elide(wrld_simpl, shift = c(k*360, 0)), col = "grey80", border = "grey80",
           add = TRUE, xlim = xlim, ylim = ylim)
  for(k in seq_along(trks)) lines(trks[[k]], col = pal[k])
}

```
Get the data from the early 2000s ready to run aniMotum
```{r, Format data from early 2000s}

tr <- read.csv('../data/raw_tracks.csv')
tr <- na.omit(tr)
colnames(tr) <- c("id","date","lat","lon","lc")
 

```
Create the plots of the raw data

```{r, Plot the raw data}
# Get the basemap data for Antarctica
bg <- ne_countries(scale = "large", country = 'united kingdom', returnclass = "sf")


p = ggplot()+

  # Sets the basemap
  geom_sf(data = bg)+
  coord_sf(xlim = range(tr$lon, na.rm = TRUE),
           ylim = range(tr$lat, na.rm = TRUE),
           expand = FALSE)+

  # Plots the tracks as lines and points, using the same colours for each individual
  geom_path(data = tr,
            aes(x=lon,y=lat,group=id,color=id),
            alpha = 1.2)+
  geom_point(data = tr,
             aes(x=lon,y=lat,group=id,fill=id),
             alpha = 0.7, shape=21, size = 2)
p <- p + ggtitle('Raw tracks')
p # Outputs the map we just created

# Save the pdf
pdf("../figures/GRSErawtracks.pdf")
p
dev.off()
```

Fit the SSM

```{r, Run aniMotum}
# Fit a SSM using a Movement Persistence model
fit.1 <- fit_ssm(tr,
                 vmax= 4,
                 model = "mp", # Use the movement persistence model
                 time.step = 2, # Get a location every 3 hours
                 control = ssm_control(verbose = 0))

fit.1.1 <- route_path(fit.1, what = "predicted")

# Extract the data for each one of the fitted models
out.1 <- grab(fit.1.1, what = "rerouted", as_sf = FALSE)

write.csv(out.1, '../results/GRSEtracks_aniMotum.csv')


```

Create the map with the tracks corrected

```{r}

m3 = ggplot()+

  # Sets the basemap
  geom_sf(data = bg)+
  coord_sf(xlim = range(out.1$lon, na.rm = TRUE),
           ylim = range(out.1$lat, na.rm = TRUE),
           expand = FALSE)+

  # Plots the tracks as lines and points, using the same colours for each individual
  geom_path(data = out.1,
            aes(x=lon,y=lat,group=id,color=id),
            alpha = 1.2)+
  geom_point(data = out.1,
             aes(x=lon,y=lat,group=id,fill=id),
             alpha = 0.7, shape=21, size = 2)

m3 <- m3 + ggtitle('Fit using a Movement Persistence model') #

m3

# Save the pdf
pdf("../figures/GESE_filteredtracks.pdf")
m3
dev.off()

```
```{r}
library(move)

# Make sure that the data are sorted by animal ID and in the correct chronological order 
# Using out.3 as that's the output from the mp model to create a move object

move.grse <- move(x=out.1$lon, 
                  y=out.1$lat,
                  time=out.1$date,
                  proj=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"), 
                  animal=out.1$id) 


# Plot the data to check we loaded the correct data
plot(move.grse, xlab="Longitude", ylab="Latitude", type="l", pch=16, lwd=0.5)


# If we wanted to check that the interpolation is every 3 hours, as we did when running the SSM, we can check as follows
steps.cese <-timeLag(move.grse, units="hours")


# Now, the next step is to calculate the distance between locations. Because the units are in degrees (lat, lon), the package move returns the distance in meters
dist.grse <- distance(move.grse)


```

Now that we have calculated some basic track metrics for the grey seal data, let's analyse the data

```{r}
# Calculate the mean step distance in meters. As these data are interpolated, then step length is giving us transit rate in km/hours
sapply(dist.grse, mean)/2000

```
To compare with the transit rate calculated by the package *move*, we can estimate the speed and mean

```{r}

speed.grse <- speed(move.grse) # Speed (or transit rate) in m/s

sapply(speed.grse, mean) # # Calculate the mean speed in meters/second
```

Now that we have estimated the transit rate in two different ways, we can compare them.


Another variable that the package *move* allows us to estimate is the heading (also known as azimuth or direction of travel/movement) of trajectory. North is represented by 0

```{r}

angle.grse <- angle(move.grse) # Calculate the heading in degrees

sapply(angle.grse, median) # Calculate the median heading in degrees
```


We need to calculate the length of the track (total distance travelled, in km)

```{r}

t.dist.cese <- sapply(dist.grse, sum) # Calculates the total distance covered by each individual, in meter

sapply(t.dist.cese, "/" , 1000) # Transforms the distance to km

```
Diving  data analysis

```{r}
library(wcUtils)

d <- list.dirs(path = "../data/", full.names = TRUE, recursive = TRUE)


for (j in (seq(2,length(d)))){
  
  counter <- j-1
  
 # Get the name of the seal
  ID <- substr(d[j],9,90)
  
  #load the ecdf file for that seal 
  
  d2 <- list.files(path = paste("../data/",ID,'/', sep = ''),pattern = 'csv')
  #
  f.name <- str_subset(d2,'-ECDHistos.csv') #Find the  file with the ecdf for that sea;
  
  s.ecdf <- read_ecdf(paste('../data/',ID,'/',f.name, sep = '')) 
  
  if (counter == 1){
    ECDF <- s.ecdf
  } else {
    ECDF <- rbind(ECDF, s.ecdf)}
  
  rm(s.ecdf)
  
  
}

ecd_data <- ECDF |> 
  dplyr::rename(deployid = deploy_id) |> 
  dplyr::arrange(deployid, start) |> 
  rowwise() |> 
  dplyr::mutate(spline_interp = list(spline_ecdf(full_ecdf, bin.width=20)),
                depth50pct = x_pct_depth(full_ecdf,pct_tad = 0.5),
                n_dives = if_else(all(is.na(n_dives_shallow),is.na(n_dives_deep)),
                                  NA, mean(c(n_dives_shallow,n_dives_deep), na.rm=TRUE))
  ) |> 
  dplyr::select(c(deployid,start:percent_dry,shallow_ecdf, 
                  deep_ecdf,full_ecdf,spline_interp,depth50pct, n_dives)) |> 
  unnest(spline_interp, keep_empty = TRUE) |> 
  group_by(deployid, start, end, kind, percent_dry) |> 
  dplyr::arrange(deployid, start, depth_break) |> 
  dplyr::mutate(propTAD = c(0,diff(ecd_prop)),
                minTAD = 60*(1-0.01*percent_dry) * propTAD) |> 
  # determine the next depth value
  dplyr::mutate(next_depth = dplyr::lead(depth_break)) |>
  # determine previous depth for plotting
  dplyr::mutate(prev_depth = dplyr::lag(depth_break)) |>
  dplyr::select(-next_depth) |>
  # transform depths to negative values
  dplyr::mutate(depth_break = depth_break * -1,
                prev_depth = prev_depth * -1)

ecd_data <- ecd_data |> 
  group_by(deployid) |> 
  arrange(start)

```

```{r}
#| fig-height: 8
#| fig-asp: 2.5
ecd_data |> 
  left_join(deploy_tbl, by = 'deployid') |> 
  dplyr::filter(between(start,
                        date_of_deployment + lubridate::hours(24),
                        data_end)) |> 
  dplyr::mutate(data_year = paste("Deployment Year:",lubridate::year(date_of_deployment))) |> 

ggplot() +
  geom_rect(aes(xmin = start, xmax = end,
                ymin = 0, ymax = percent_dry/10),
            fill = "seashell3") +
  geom_rect(aes(xmin = start, xmax = end,
                ymin = depth_break, ymax = prev_depth,
                fill = propTAD),
            color = NA) +
  scale_fill_distiller(palette = "Reds",
                   direction = 1, trans = "log10",
                       guide = guide_colorbar(
                         title = 'proportion of time submerged',
                         title.position = 'bottom',
                         title.hjust = 0.5,
                         barwidth = unit(75, units = "mm"),
                         barheight = unit(2, units = "mm"))) +
  geom_smooth(aes(y = depth50pct*-1, x = start), se=FALSE,
              linewidth = 0.2, na.rm=TRUE, color = "black",
              
              method = "gam", formula = y ~ s(x, bs = "cs", k=24)) +
  facet_wrap(deployid ~ data_year, scales = "free_x",ncol=2) +
  ylab("depth (meters)") +
  labs(title = "Grey Seals - Proportion of Time at Depth",
       subtitle = stringr::str_wrap("Each vertical bar represents 1 hour binned at 20m 
                                    depth increments. The black line represents a 
                                    smoothed GAM fit through the 50% depth break."),
       caption = stringr::str_wrap("data derived from the Emperical Cummulative 
                                   Distribution (ECD) of time at depth 
                                   transmitted via the Argos satellite network")) +
  theme_minimal() +
  theme(legend.position = "bottom")




```
```{r}

carto_blue <- rcartocolor::carto_pal(n=5,name="Bold")[3]
#| fig-asp: 1.2
ecd_data |> 
  left_join(deploy_tbl, by = 'deployid') |> 
  dplyr::filter(between(start,
                        date_of_deployment + lubridate::hours(24),
                        data_end),
                depth_break < 0,
                !is.na(depth_break)) |> 
  dplyr::mutate(depth_break = factor(depth_break,levels = seq(0,-480,by=-20))) |> 
  ggplot(aes(x=propTAD, y = depth_break)) +
  # geom_boxplot(aes(group = depth_break), varwidth = TRUE,
  #              outlier.alpha = 0.2) +
  ggdist::stat_gradientinterval(
    color = carto_blue, fill = carto_blue,
    point_color = "black",
    normalize = "groups",
    na.rm=TRUE
  ) +
  scale_y_discrete(limits=rev) +
  ylab("depth bin (lower bound)") +
  xlab("proportion of time at depth") +
  labs(title = "Grey Seals - Summarized Time at Depth",
       subtitle = stringr::str_wrap("Each horizontal strip represents the distribution
       of proportional time at depth for each 20m depth increment. Points indicate
       the median value."),
       caption = stringr::str_wrap("data derived from the Emperical Cummulative 
                                   Distribution (ECD) of time at depth 
                                   transmitted via the Argos satellite network")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

